Modal Verification Patterns for Systems Software

### 2025-09-13 paper.tex
+

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{On Verification Patterns for Low-Level Systems via Modal Abstractions}

@@

\textit{Contributions} We argue that modal abstractions can be used to identify and abstract system verification challenges. We justify our perspective by discussing prior systems that have successfully used modalities for system verification, arguing that they fit into the verification design pattern we articulate, and explaining how this approach might apply to other systems' verification challenges.
\begin{itemize}
\item \textbf{Identifying System Verification Challenges}: We start with identifying common patterns in system verification: \emph{virtualization, sharing, and translation}.
\item \textbf{Introducing Modal \emph{Resource Contexts}}: Then we discuss the concept of \emph{resource} which has already been an essential concept in the design of systems. Inspired by the concept of resource in systems, we define what a resource and its context are in the modal abstractions. For example, the \emph{transaction} abstraction, which behaves like a container for in-memory dirty disk pages, can be considered as a resrouce context for any dirty in-memory disk page resource associated with it.
\item \textbf{Introducing \emph{Nominals for System Resources}}:
Existing logics for systems software must frequently identify
individual resources within a given context ---
nominalization (in the sense of hybrid logic) enables identifying a resource in its context. 
For example, a filesystem transaction is a context of resources for in-memory updated disk blocks
which have not yet been saved to disk.
The transaction identifier is used to associate a transaction with a disk-block to be persisted so that, 
in case of a crash while persisting updated disk-blocks, the filesystem can rollback the 
\emph{already persisted} disk-blocks of the transaction, and reach to the previous consistent disk state. 
We discuss other examples to contextualize the existing systems surveyed.
% to be able to do so, both the updated in-memory disk blocks and the transaction must refer to the transaction 
% identifier -- \emph{strong nominalization}. 
% in another example, virtual memory references (resources) in 
% an address space (resource context), which can be uniquely identified with a root address (the nominal) 
% of its page-table tree, are \emph{agnostic} to the address space that they are in. 
% However, they can only be accessed (be valid) in the address space to which they are agnostic. 
% However, then an address space switch happens, the virtual memory references of the previous address 
% space must be made inaccessible. To be able to do so, although virtual memory references do not hold any 
% piece of information related to their address space, we still have to associate them. 
% We call this kind of unilateral nominalization -- \emph{weak nominalization}.
% \todo[inline,color=cyan]{I left this paragraph alone, mostly, but it's too detailed for the new intro section position here}
\item \textbf{Taxonomy of Current Modal Approaches in System Verification}: Based on these concepts defined, we summarize contemporary verification efforts using modal abstractions. We choose them from different domains, for example, reasoning about weak memory models and storage persistence ~\cite{tejthesis,perennialgit,tejperennial19,larsnextgen25,fsl,fsl++,derekrustbelt20,kuru2025modalabstractionsvirtualizingmemory,amalreal2024} because we would like to justify that our \emph{definitions} are not domain dependent.
\end{itemize}

@@

Conventionally, a Kripke model with its constituent must be picked. Gordon give a Kripke model for an actor-based hybrid modal logic ~\cite{gordon2019modal} in which the transition relation is indexed by a particular command and a particular actor. Wagner et.al. takes the similar approach for the design of their program logic for reference counted memory.

@@

\paragraph{Reference Counting} When dealing with reference counting APIs, we may care to specify reachability of memory nodes ($\ell \mapsto v$) in a certain context defined by a shared root address. A shared memory address $\ell$ can be the root of the graph ($@_{\ell}\; ( \ell_1 \mapsto v_1 \;\ast\; ... \; \ast \; \ell_j \mapsto v_j )$) that can be a container of memory nodes ($\ell_1 \mapsto v$) that are reachable from the root $\ell$.

\paragraph{Post-Crash}
%\Colin{Ismail, not sure how to describe this. We don't want to go into the weeds on the NextGen}
%that can be jumped from a shared memory referenced whose reference count is at least 1 are also example resource elements from memory management task at different abstraction levels. 
The resource element of Post-Crash Modality is not obvious in Table \ref{table:decomposition}, and needs a bit of explanation. Perennial, based on the Iris logic, has both disk-points-to assertions $d[\textsf{p}] \mapsto_{n} v$ (for a specific disk $d$) and in-memory points-to assertions $\ell \mapsto_{n}^{\overline{\gamma}}$. Perennial crash-recovery logic book-keeps resource names (can be thought of as logical variables) $\overline{\gamma}$ to identify 
which assertions (resource elements) remain valid after a crash --- these assertions
are only usable while the names in $\overline{\gamma}$ are valid, and a crash resets
them, discarding assumptions about volatile state. A subtlety of the notion of a resource context is that, unlike the earlier examples, the context does not need to be a literal data structure. It can instead be (various forms of) a set of executions, as in the Post-Crash and NextGen modalities.
The Post-Crash modality $\diamond$ expresses that the assertion $P$ will be true after a crash discards
all unstable storage (i.e., RAM).
This was the inspiration for the NextGen modality, which is in fact a framework for defining ``after-$t$'' modalities, where $t$ is an transformation of the global state.\footnote{The transformations are subject to some
technical constraints that are unrelated to our point here.}
%whose capability is considered to be relevant to be kept in the crash invariant $\fbox{C}_n$, and to be used when a crash occurs.\Colin{Ismail, not sure about last line above. Can you better explain gamma?}
%regions of state (including auxiliary/ghost state) which persist after a crash (anything not identified

>>>>

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Modal Verification Patterns for Systems Software}

@@

\textit{Contributions} We argue that modal abstractions can be used to identify and abstract system verification challenges. We justify our perspective by discussing prior systems that have successfully used modalities for system verification, arguing that they fit into the verification design pattern we articulate, and explaining how this approach might apply to other systems' verification challenges.
\begin{itemize}
\item \textbf{Identifying System Verification Challenges}: We start with identifying common patterns in system verification: \emph{virtualization, sharing, and translation}.
\item \textbf{Introducing Modal \emph{Resource Contexts}}: Then we discuss the concept of \emph{resource} which has already been an essential concept in the design of systems. Inspired by the concept of resource in systems, we define what a resource and its context are in the modal abstractions. For example, the \emph{transaction} abstraction, which behaves like a container for in-memory dirty disk pages, can be considered as a resrouce context for any dirty in-memory disk page resource associated with it.
\item \textbf{Introducing \emph{Nominals for System Resources}}:
Existing logics for systems software must frequently identify
individual resources within a given context ---
nominalization (in the sense of hybrid logic) enables identifying a resource in its context. 
For example, a filesystem transaction is a context of resources for in-memory updated disk blocks
which have not yet been saved to disk.
The transaction identifier is used to associate a transaction with a disk-block to be persisted so that, 
in case of a crash while persisting updated disk-blocks, the filesystem can rollback the 
\emph{already persisted} disk-blocks of the transaction, and reach to the previous consistent disk state. 
We discuss other examples to contextualize the existing systems surveyed.
% to be able to do so, both the updated in-memory disk blocks and the transaction must refer to the transaction 
% identifier -- \emph{strong nominalization}. 
% in another example, virtual memory references (resources) in 
% an address space (resource context), which can be uniquely identified with a root address (the nominal) 
% of its page-table tree, are \emph{agnostic} to the address space that they are in. 
% However, they can only be accessed (be valid) in the address space to which they are agnostic. 
% However, then an address space switch happens, the virtual memory references of the previous address 
% space must be made inaccessible. To be able to do so, although virtual memory references do not hold any 
% piece of information related to their address space, we still have to associate them. 
% We call this kind of unilateral nominalization -- \emph{weak nominalization}.
% \todo[inline,color=cyan]{I left this paragraph alone, mostly, but it's too detailed for the new intro section position here}
\item \textbf{Taxonomy of Current Modal Approaches in System Verification}: Based on these concepts defined, we summarize contemporary verification efforts using modal abstractions. We choose them from different domains, for example, reasoning about weak memory models and storage persistence ~\cite{tejthesis,perennialgit,tejperennial19,larsnextgen25,fsl,fsl++,derekrustbelt20,kuru2025modalabstractionsvirtualizingmemory,amalreal2024,gordon2019modal} and using different programming logics such as Iris ~\cite{jung2018iris} and Dafny ~\cite{leino2010dafny} because we would like to justify that our \emph{definitions} are neither domain nor programming logic dependent. 
\end{itemize}

@@

Conventionally, a Kripke model with its constituent must be picked. Gordon give a Kripke model for an actor-based hybrid modal logic ~\cite{gordon2019modal} in which the transition relation is indexed by a particular command and a particular actor which relates pre- and post-states of a particular actor referenced by a nominal executing the command. Gordon encodes this hybrid logic in Dafny program logic ~\cite{leino2010dafny} with the help of two state invariants and rely-guarantee assertions checking the stability of invariants. Wagner et.al. takes the similar approach for the design of their separation logic and pick up a Kripke model for reference counted memory reasoning principles.

@@

\paragraph{Reference Counting} When dealing with reference counting APIs, we may care to specify reachability of memory nodes ($\ell \mapsto v$) in a certain context defined by a shared root address. A shared memory address $\ell$ can be the root of the graph ($@_{\ell}\; ( \ell_1 \mapsto v_1 \;\ast\; ... \; \ast \; \ell_j \mapsto v_j )$) that can be a container of memory nodes ($\ell_1 \mapsto v$) that are reachable from the root $\ell$. 
\paragraph{Actors} When dealing with reference counting APIs, we may care to specify reachability of memory nodes ($\ell \mapsto v$) in a certain context defined by a shared root address. A shared memory address $\ell$ can be the root of the graph ($@_{\ell}\; ( \ell_1 \mapsto v_1 \;\ast\; ... \; \ast \; \ell_j \mapsto v_j )$) that can be a container of memory nodes ($\ell_1 \mapsto v$) that are reachable from the root $\ell$. 
\paragraph{Post-Crash}
%\Colin{Ismail, not sure how to describe this. We don't want to go into the weeds on the NextGen}
%that can be jumped from a shared memory referenced whose reference count is at least 1 are also example resource elements from memory management task at different abstraction levels. 
The resource element of Post-Crash Modality is not obvious in Table \ref{table:decomposition}, and needs a bit of explanation. Perennial, based on the Iris logic, has both disk-points-to assertions $d[\textsf{p}] \mapsto_{n} v$ (for a specific disk $d$) and in-memory points-to assertions $\ell \mapsto_{n}^{\overline{\gamma}}$. Perennial crash-recovery logic book-keeps resource names (can be thought of as logical variables) $\overline{\gamma}$ to identify 
which assertions (resource elements) remain valid after a crash --- these assertions
are only usable while the names in $\overline{\gamma}$ are valid, and a crash resets
them, discarding assumptions about volatile state. A subtlety of the notion of a resource context is that, unlike the earlier examples, the context does not need to be a literal data structure. It can instead be (various forms of) a set of executions, as in the Post-Crash and NextGen modalities.
The Post-Crash modality $\diamond$ expresses that the assertion $P$ will be true after a crash discards
all unstable storage (i.e., RAM).
This was the inspiration for the NextGen modality, which is in fact a framework for defining ``after-$t$'' modalities, where $t$ is an transformation of the global state.\footnote{The transformations are subject to some
technical constraints that are unrelated to our point here.}
%whose capability is considered to be relevant to be kept in the crash invariant $\fbox{C}_n$, and to be used when a crash occurs.\Colin{Ismail, not sure about last line above. Can you better explain gamma?}
%regions of state (including auxiliary/ghost state) which persist after a crash (anything not identified
